Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-08-15 manuscript/manuscript.Rmd
Final recompilation and ce for Frontiers

```{r supp table 1 standards list}
clean_stans <- read_csv("../metadata/clean_stans.csv")
raafay_notes <- read_csv("../targeted/raw_data/Raafay_HILIC_notes.csv") %>%
  select(compound_name, in_samps)

IS_table <- clean_stans %>%
  filter(compound_type=="Internal Standard") %>%
  select(compound_name, formula, stan_mz, stan_rt, conc_um, date_added, polarity) %>%
  mutate(ms_run=case_when(
    date_added<180205~"All",
    date_added<=180821~"MESO-SCOPE Eddy Transect and HEE",
    date_added<=190715~"HEE only",
    TRUE~"None"
  )) %>%
  filter(ms_run!="None") %>%
  mutate(polarity=ifelse(polarity=="pos", "Positive", "Negative")) %>% 
  mutate(stan_type=case_when(
    compound_name%in%c("L-Cysteic acid, 2H3", "3-Sulfolactate, 13C3", "Sulfoacetic acid, 13C2", "Isethionic acid, 13C2", "Taurine, 2H4")~"Extraction",
    TRUE~"Internal"
  )) %>%
  arrange(stan_mz) %>%
  select(`Compound name`=compound_name, `Molecular formula`=formula,
         `m/z ratio`=stan_mz, `Retention time (min)`=stan_rt,
         `Concentration (uM)`=conc_um, `Polarity integrated`=polarity,
         `Which runs`=ms_run, `Extraction or internal`=stan_type)

Auth_table <- clean_stans %>%
  mutate(ms_run=case_when(
    date_added<180205~"All",
    date_added<=180821~"MESO-SCOPE Eddy Transect and HEE",
    date_added<=190715~"HEE only",
    TRUE~"None"
  )) %>%
  filter(ms_run!="None") %>%
  left_join(raafay_notes) %>%
  mutate(polarity=ifelse(polarity=="pos", "Positive", "Negative")) %>%
  arrange(stan_mz) %>%
  select(`Compound name`=compound_name, `Molecular formula`=formula,
         `m/z ratio`=stan_mz, `Retention time (min)`=stan_rt,
         `Concentration (uM)`=conc_um, `Polarity integrated`=polarity,
         `Standard mix`=mix, `Which runs`=ms_run, `Good quality`=in_samps)

@@

```{r PC_vs_totnM regression values, include=FALSE}
frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity, n_C, n_N)) %>%
  mutate(nM_C=n_C*nM, nM_N=n_N*nM) %>%
  group_by(filename) %>%
  summarize(total_nM_C=sum(nM_C), total_nM_N=sum(nM_N)) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(frac_C_in_metab=total_nM_C/1000/PC_um,
         frac_N_in_metab=total_nM_N/1000/PN_um) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  ungroup()

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a relatively fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 2). Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

@@

The most abundant intracellular metabolite quantified in the eddy transect samples was trimethylamine N-oxide, with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 2), though a few samples had concentrations an order of magnitude higher for unknown reasons (Supplemental Figure 3). A majority of the known molecules (37/53) had a similar pattern with a subsurface maximum at the DCM, including glycine betaine, glutamate, and guanine. Thirteen molecules obeyed a different pattern and decreased monotonically with depth, including the molecules hydroxyisoleucine, gonyol, and dimethylsulfoniopropionate. Two known molecules (arsenobetaine and O-acetylcarnitine) increased with depth, potentially representing particle degradation during export.

@@

library(arrow)
dset <- open_dataset("../tmzMLs/MT_all_file_arr_pqds/")
arrow_data <- ms11_df %>%
  pmap(function(...){
    row_data <- list(...)
    dset %>%
      filter(mz%between%pmppm(row_data$mzmed, 20)) %>%
      collect() %>%
      mutate(formula=row_data$formula)
  }, .progress = TRUE)
Nacetyl_row <- ms11_df %>% filter(formula=="C4H10NO4S")
neg_msdata <- grabMSdata(list.files("../tmzMLs/neg", pattern = "180821_Smp", full.names = TRUE))
Nacetyldata <- neg_msdata$MS1[mz%between%pmppm(Nacetyl_row$mzmed-1.007276*2, 10)] %>%
  mutate(formula=Nacetyl_row$formula) %>%
  mutate(filename=str_replace(filename, "tmzML", "mzML"))
peak_lims <- arrow_data %>%
  bind_rows() %>%
  filter(str_detect(filename, "Smp")) %>%
  bind_rows(Nacetyldata) %>%
  distinct() %>%
  left_join(ms11_df) %>%
  filter(rt>rtmed-1 & rt<rtmed+1)

@@

mass_diffs <- peak_lims %>%
  group_by(formula, rtmed, mzmed, calc_mass) %>%
  summarise(mean_mz=weighted.mean(mz, int)) %>%
  mutate(calc_mass=ifelse(formula=="C4H10NO4S", calc_mass-2*1.007276, calc_mass)) %>%
  mutate(delta_ppm=round((mean_mz-calc_mass)*1e6/mzmed, 2)) %>%
  select(-calc_mass)
ms11_df %>%
  left_join(mass_diffs) %>%
  select(`Median m/z`=mzmed, `Median RT (min)`=rtmed, `Putative formula`=formula,
         `Delta m/z (ppm)`=delta_ppm, `Compound name`=compound_name) %>%
  knitr::kable()
```

@@

```{r trichobloomsamples comparison, eval=FALSE}
library(RaMS)
trichodata <- list.files("../mzMLs/tricho_pos", pattern=".mzML", full.names = TRUE) %>%
  grabMSdata()
tricho_eics <- ms11_df %>%
  pmap(function(...){
  row_data <- list(...)
  peak_data <- trichodata$MS1[mz%between%pmppm(row_data$mzmed, 10)]
  cbind(peak_data, formula=row_data$formula)
}) %>%
  bind_rows()
tricho_eics %>%
  filter(rt%between%c(2, 15)) %>%
  arrange(mz) %>%
  mutate(formula=fct_inorder(formula)) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename)) +
  facet_wrap(~formula, scales="free", ncol=2) +
  geom_vline(aes(xintercept = rtmed), color="red", data=ms11_df)
trichodata$MS1[mz%between%pmppm(162.1125, 10)] %>%
  mutate(samp_type=str_extract(filename, "Smp|Poo|Mix1|Mix2|Mat")) %>%
  filter(samp_type!="Mat") %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=samp_type)) +
  ggtitle("Homoserine betaine in Tricho bloom samples")
```

@@

Given the confluence of both depth and SLA signals and the way that the boxplots of Figure 2 would confound the SLA signal with the depth variance, we used k-means clustering to group similar compounds and provide a reduced dimensionality space for visualization. This revealed four major patterns of metabolite response, with a majority of compounds responding to eddy state (Clusters 1, 2, and 4 in Figure 5). It is interesting to note that the sole cluster in which abundances in the anticyclone are greater than those of the cyclone is also the only cluster to generally increase in abundance with depth (Cluster 4), perhaps indicating that compounds more concentrated in the anticyclone are the same kind of degradation products and recalcitrant carbon typically found at depth.

>>>>

```{r supp table 1 standards list}
clean_stans_supp <- read_csv("../metadata/clean_stans.csv")
raafay_notes <- read_csv("../targeted/raw_data/Raafay_HILIC_notes.csv") %>%
  select(compound_name, in_samps)

IS_table <- clean_stans_supp %>%
  filter(compound_type=="Internal Standard") %>%
  select(compound_name, formula, stan_mz, stan_rt, conc_um, date_added, polarity) %>%
  mutate(ms_run=case_when(
    date_added<180205~"All",
    date_added<=180821~"MESO-SCOPE Eddy Transect and HEE",
    date_added<=190715~"HEE only",
    TRUE~"None"
  )) %>%
  filter(ms_run!="None") %>%
  mutate(polarity=ifelse(polarity=="pos", "Positive", "Negative")) %>% 
  mutate(stan_type=case_when(
    compound_name%in%c("L-Cysteic acid, 2H3", "3-Sulfolactate, 13C3", "Sulfoacetic acid, 13C2", "Isethionic acid, 13C2", "Taurine, 2H4")~"Extraction",
    TRUE~"Internal"
  )) %>%
  arrange(stan_mz) %>%
  select(`Compound name`=compound_name, `Molecular formula`=formula,
         `m/z ratio`=stan_mz, `Retention time (min)`=stan_rt,
         `Concentration (uM)`=conc_um, `Polarity integrated`=polarity,
         `Which runs`=ms_run, `Extraction or internal`=stan_type)

Auth_table <- clean_stans_supp %>%
  mutate(ms_run=case_when(
    date_added<180205~"All",
    date_added<=180821~"MESO-SCOPE Eddy Transect and HEE",
    date_added<=190715~"HEE only",
    TRUE~"None"
  )) %>%
  filter(ms_run!="None") %>%
  left_join(raafay_notes) %>%
  mutate(polarity=ifelse(polarity=="pos", "Positive", "Negative")) %>%
  arrange(stan_mz) %>%
  select(`Compound name`=compound_name, `Molecular formula`=formula,
         `m/z ratio`=stan_mz, `Retention time (min)`=stan_rt,
         `Concentration (uM)`=conc_um, `Polarity integrated`=polarity,
         `Standard mix`=mix, `Which runs`=ms_run, `Good quality`=in_samps)

@@

```{r frac_pc_in_metab_df creation}
frac_pc_in_metab_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity, n_C, n_N)) %>%
  mutate(nM_C=n_C*nM, nM_N=n_N*nM) %>%
  group_by(filename) %>%
  summarize(total_nM_C=sum(nM_C), total_nM_N=sum(nM_N)) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(frac_C_in_metab=total_nM_C/1000/PC_um,
         frac_N_in_metab=total_nM_N/1000/PN_um) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))
```

```{r PC_vs_totnM regression values, include=FALSE}
no_outlier_frac_pc_in_metab_df <- frac_pc_in_metab_df %>%
  filter(!(station=="11" & tripl=="A" & depth=="Surface")) %>%
  ungroup()

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental Figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental Figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a relatively fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 3). Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

@@

The most abundant intracellular metabolite quantified in the eddy transect samples was trimethylamine N-oxide, with an average concentration of 1.38 nM and a distinct DCM maximum (Figure 3), though a few samples had concentrations an order of magnitude higher for unknown reasons (Supplemental Figure 3). A majority of the known molecules (37/53) had a similar pattern with a subsurface maximum at the DCM, including glycine betaine, glutamate, and guanine. Thirteen molecules obeyed a different pattern and decreased monotonically with depth, including the molecules hydroxyisoleucine, gonyol, and dimethylsulfoniopropionate. Two known molecules (arsenobetaine and O-acetylcarnitine) increased with depth, potentially representing particle degradation during export.

@@

library(arrow)
dset <- open_dataset("../tmzMLs/MT_all_file_arr_pqds/")
arrow_data <- ms11_df %>%
  pmap(function(...){
    row_data <- list(...)
    dset %>%
      filter(mz%between%pmppm(row_data$mzmed, 20)) %>%
      collect() %>%
      mutate(formula=row_data$formula)
  })
Nacetyl_row <- ms11_df %>% filter(formula=="C4H10NO4S")
neg_msdata <- grabMSdata(list.files("../tmzMLs/neg", pattern = "180821_Smp", full.names = TRUE), verbosity = 0)
Nacetyldata <- neg_msdata$MS1[mz%between%pmppm(Nacetyl_row$mzmed-1.007276*2, 10)] %>%
  mutate(formula=Nacetyl_row$formula) %>%
  mutate(filename=str_replace(filename, "tmzML", "mzML"))
peak_lims <- arrow_data %>%
  bind_rows() %>%
  filter(str_detect(filename, "Smp")) %>%
  bind_rows(Nacetyldata) %>%
  distinct() %>%
  left_join(ms11_df) %>%
  filter(rt>rtmed-1 & rt<rtmed+1)

@@

mass_diffs <- peak_lims %>%
  group_by(formula, rtmed, mzmed, calc_mass) %>%
  summarise(mean_mz=weighted.mean(mz, int)) %>%
  mutate(calc_mass=ifelse(formula=="C4H10NO4S", calc_mass-2*1.007276, calc_mass)) %>%
  mutate(delta_ppm=round((mean_mz-calc_mass)*1e6/mzmed, 2)) %>%
  select(-calc_mass)
final_ms11_df <- ms11_df %>%
  left_join(mass_diffs) %>%
  select(`Median m/z`=mzmed, `Median RT (min)`=rtmed, `Putative formula`=formula,
         `Delta m/z (ppm)`=delta_ppm, `Compound name`=compound_name)
# knitr::kable(final_ms11_df)
```

@@

```{r trichobloomsamples comparison, eval=FALSE}
library(RaMS)
trichodata <- list.files("../mzMLs/tricho_pos", pattern=".mzML", full.names = TRUE) %>%
  grabMSdata(verbosity = 0)
tricho_eics <- ms11_df %>%
  pmap(function(...){
  row_data <- list(...)
  peak_data <- trichodata$MS1[mz%between%pmppm(row_data$mzmed, 10)]
  cbind(peak_data, formula=row_data$formula)
}) %>%
  bind_rows()
tricho_eics %>%
  filter(rt%between%c(2, 15)) %>%
  arrange(mz) %>%
  mutate(formula=fct_inorder(formula)) %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename)) +
  facet_wrap(~formula, scales="free", ncol=2) +
  geom_vline(aes(xintercept = rtmed), color="red", data=ms11_df)
trichodata$MS1[mz%between%pmppm(162.1125, 10)] %>%
  mutate(samp_type=str_extract(filename, "Smp|Poo|Mix1|Mix2|Mat")) %>%
  filter(samp_type!="Mat") %>%
  ggplot() +
  geom_line(aes(x=rt, y=int, group=filename, color=samp_type)) +
  ggtitle("Homoserine betaine in Tricho bloom samples")
```

@@

Given the confluence of both depth and SLA signals and the way that the median metabolite plots of Figure 2 would confound the SLA signal with the depth variance, we used k-means clustering to group similar compounds and provide a reduced dimensionality space for visualization. This revealed four major patterns of metabolite response, with a majority of compounds responding to eddy state (Clusters 1, 2, and 4 in Figure 5). It is interesting to note that the sole cluster in which abundances in the anticyclone are greater than those of the cyclone is also the only cluster to generally increase in abundance with depth (Cluster 4), perhaps indicating that compounds more concentrated in the anticyclone are the same kind of degradation products and recalcitrant carbon typically found at depth.
